Effects of epidemiological structure on the transient evolution of HIV virulence

### 2017-01-12 ms/ms.tex
changing to lspvl...

Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of the expected time of progression
to AIDS (\figurename~\ref{fig:virtraj}: lower progression time corresponds
to higher virulence). Although the speed of virulence evolution varies, leading to wide variation in the peak SPVL (ranging from \todo{XX} to \todo{YY})), virulence peaks in all models between 200 and 300 years.

@@

Our chosen summary statistics (peak time, minimum expected progression time, equilibrium expected
progression time, and relative progression time) all vary considerably across models
(\figurename~\ref{fig:unidist}).
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. Some parameter
sets for these models lead to low equilibrium virulence ($\approx 18$ years
to progression);
these same sets lead to correspondingly low
peak virulence (16 years to progression) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
At the opposite extreme, parameter sets that produce high equilibrium virulence (8 years to progression)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (4 years to progression).
The pair-formation without extra-pair contact and implicit models
occasionally have parameter sets that select for such low virulence across
the board that they never exceed their initial virulence, leading to a tail
of peak times near zero.

@@

These differences are practically as well as 
scientifically important. The random-mixing, pairform+epc,
and heterogeneous models all predict rapid
progression to AIDS at the virulence peak
(median/95\% CI = 6.1 (5.7-6.3), 
6.02 (5.04-7.7), 6.03  (4.8-9.2)). 
In contrast, 
the implicit model predicts minimum progression times about twice as long:
12.5 (9.6-15.6) years. The corresponding differences in 
within-couple transmission
probability are even more extreme, about a fourfold difference:
0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
random and pairform+epc models vs. 0.059 (0.02-0.13) per year
for the implicit model (see Appendix S2 for 
plots showing univariate summaries
of \Lspvl\ and transmission probability).

@@

The bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
different models with similar univariate dynamical summaries. While the
relationship between equilibrium progression time and peak time is
similar for all model structures (top left panel), the other
relationships show more variation. In particular, the implicit
and pair-formation (without extra-pair contact) are very similar
to each other, but distinct from the other models. We still do
not have a convincing explanation for this distinction; we
would have expected the implicit model to be most similar to the
the instantaneous-switching model without extra-pair contact,
which most closely matches its derivation. However, we note
that the implicit model derivation is based on defining
the force of infection to match a scaled version of $\rzero$,
and as such would be expected to match the equilibrium behaviour
but not necessarily the epidemic-phase behaviour of a model
with explicit partnership dynamics.

@@

In the particular example of HIV virulence eco-evolutionary dynamics
and the complexity of contact structures
we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, putting us back where we started when used a naive random-mixing
contact model.
However, we are not quite back where we started, as the
complex models lead to wider, presumably more realistic
confidence intervals on the predictions.
In general, unstructured mixing --- whether occurring through 
purely random mixing, or through extra-pair contact and contact
among people outside of stable partnerships --- tends to drive
faster virulence evolution, leading to higher peak virulence and 
lower times to progression at the peak time.

>>>>

Across the entire range of parameters covered by the LHS analysis, all of the classes of models we considered produce qualitatively similar virulence trajectories, which we quantify in terms of population mean \Lspvl\ (\figurename~\ref{fig:virtraj}: higher population mean \Lspvl\ corresponds
to higher virulence). Although the speed of virulence evolution varies, leading to wide variation in the peak SPVL (ranging from 3 to 5.5), virulence peaks in all models between 200 and 300 years.

@@

Our chosen summary statistics (peak time, maximum mean \Lspvl, equilibrium mean \Lspvl, and minimum mean progression time) all vary considerably across models
(\figurename~\ref{fig:unidist}).
We first consider the models of intermediate realism: implicit,
instantaneous-switching with and without extra-pair contact, and
pair formation without extra-pair contact. Some parameter
sets for these models lead to low equilibrium virulence ($\approx 2.3$ \Lspvl);
these same sets lead to correspondingly low
peak virulence (\Lspvl\ of 3 \todo{This is same as the initial value... how did you decide on this value in terms of progression time before? Peak time is 0 in this case...}) and early peak times (before 200 years: \figurename~\ref{fig:pairplot}).
At the opposite extreme, parameter sets that produce high equilibrium virulence (4.7 \Lspvl)
also produce late peaks ($> 200~\text{years}$) and
high peak virulence (5.6 \Lspvl).
The pair-formation without extra-pair contact and implicit models
occasionally have parameter sets that select for such low virulence across
the board that they never exceed their initial virulence, leading to a tail
of peak times near zero \todo{and also implying that we can observe a peak virulence lower than 3 \Lspvl\ if we start with a lower initial condition}.

@@

These differences are practically as well as 
scientifically important. The random-mixing, pairform+epc,
and heterogeneous models all predict high population mean \Lspvl\ at the virulence peak
(median/95\% CI = 4.7 (4.65-4.79), 
4.72 (4.37-4.96), 4.72  (4.09-5.03)). 
In contrast, 
the implicit model predicts a much lower peak \Lspvl\ value:
3.52 (3-4.02) years.
\todo{I'm not sure if you want to leave the transmission rates here or not...}
\todo{I guess we can keep the progression time here because we have it in our plot?}
The corresponding differences in 
within-couple transmission
probability are even more extreme, about a fourfold difference:
0.249 (0.24-0.26), 0.252 (0.19-0.28), and 0.252 (0.15-0.28) per year for the 
random and pairform+epc models vs. 0.059 (0.02-0.13) per year
for the implicit model (see Appendix S2 for 
plots showing univariate summaries
of expected progression time to AIDS and transmission probability).

@@

The bivariate relationships (\figurename~\ref{fig:pairplot}) help distinguish the results of 
different models with similar univariate dynamical summaries. While the
relationship between equilibrium \Lspvl\ and peak time is
similar for all model structures (top left panel), the other
relationships show more variation. In particular, the implicit
and pair-formation (without extra-pair contact) are very similar
to each other, but distinct from the other models. We still do
not have a convincing explanation for this distinction; we
would have expected the implicit model to be most similar to the
the instantaneous-switching model without extra-pair contact,
which most closely matches its derivation. However, we note
that the implicit model derivation is based on defining
the force of infection to match a scaled version of $\rzero$,
and as such would be expected to match the equilibrium behaviour
but not necessarily the epidemic-phase behaviour of a model
with explicit partnership dynamics.

@@

In the particular example of HIV virulence eco-evolutionary dynamics
and the complexity of contact structures
we reach the slightly ironic conclusion that the
effort put into building a more realistic model essentially cancels
out, putting us back where we started when used a naive random-mixing
contact model.
However, we are not quite back where we started, as the
complex models lead to wider, presumably more realistic
confidence intervals on the predictions.
In general, unstructured mixing --- whether occurring through 
purely random mixing, or through extra-pair contact and contact
among people outside of stable partnerships --- tends to drive
faster virulence evolution, leading to higher peak virulence and \todo{I'm guessing you still want to keep progression time here?}
lower times to progression at the peak time.
